Existence of Forward Simulations for Particular Data Structures

### 2017-01-23 draft.tex
TS Stack linearizability proof, push cases finished

void push(int x) {
  Node* n = new Node(x,MAX_INT,
                        null,false);
  n->next = pools[myTID];
  pools[myTID] = n;
  int i = TS++;
  n->ts = i;
}
int pop() {
 boolean success = false;
 int maxTS = -1;
 Node* youngest = null;
 while ( !success ) {
   maxTS = -1; youngest = null;
   for(int i=0; i<maxThreads; i++){
     Node* n = pools[i];
     while (n->taken && n->next != n)
       n = n->next;
     if(maxTS < n->ts) {
       maxTS = n->ts; youngest = n;
     }
   }
   if (youngest != null)
     success=CAS(youngest->taken,
                       false,true);
 }
 return youngest->data;
}
\end{lstlisting}
Then, we begin to obtain the LTS ($L_C$) of TS Stack from the algorithm by defining control points and the actions among these control points as seen in Figure X \textcolor{red}{Flow diagram is referenced here}. To simplify the proof, we take the initializations of some local variables together as atomic.

States of the TS-Stack contains the global variables and local variables as fields. Global variables are just elements of their domains and local variables are maps from operation identifiers to their domains. We say $i_q(k)$ for referencing the value of local variable $i$ of operation $k$ in state $q$. There is only one special local variable called $myTID$. Its value is unique to each pending operation in a state i.e., concurrent operations cannot have the same $myTID$ value. TS-Stack states also contains sets $O_a, O_r \in \mathbb{O}$ which are operation identifier sets of push and pops respectively, and the control point function $cp$ which is a map from operation identifiers to the control points set that are presented in the flow diagram Figure X  \textcolor{red}{Flow diagram is referenced here}. Transition relation of the TS-STack is presented in Figure~\ref{fig:transitions:TSS}.
Next, we show that the linearizability of TS Stack.

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp) \\ 
      d\neq {\tt EMPTY} \\
      \forall k'.\ ov'(k') = ov(k')\cup \{k\}
    }{
      O,<,\ell,rv,cp,be,ov 
      \xrightarrow{inv(push,d,k)} 
      O\cup\{k\},<\cup\ {\tt COMP}(O)\times\{k\},\ell[k\mapsto (d,{\tt PEND})],rv,cp[k\mapsto A_1],be,ov'
    }\hspace{5mm}

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The transition relation of TS-Stack
  }
  \label{fig:transitions:TSS}
\vspace{-6mm}
\end{figure}

\begin{lem}
$L_C$ is a Com(pop)- refinement of $AbsS$.
\end{lem}
\begin{proof}
Since $AbsS$ is deterministic with respect to $C \cup R \cup Com(pop)$, there exists a Com(pop) forward simulation from $L_C$ to $AbsS$ iff $L_C$ is a Com(pop)-refinement of $AbsS$. Hence, we define the relation $fs$ and show that it is a Com(pop)-forward simulation relation.

@@

\item[\emph{TraverseBefore}] If a pop operation $p$ is currently visiting node $n$ and there is a node $m$ coming before $n$ in the traverse order with a greater timestamp, then the operation that inserts $m$ must be concurrent with $p$. More formally, let $k \in O_t$ such that $n_s(k) <^{tr}_s n_s(p)$ and $n_s(k)->ts > n_s(p)->ts$. Then, $k \in ov_t(p)$.
\item[\emph{TraverseAfter}] If a pop operation $p$ is currently visiting node $n$ that is not null and its youngest element $m$ is not null and still not taken in state $s$, then either $m$ is a candidate to be removed by $p$ in $t$ or there exists a later node $m'$ than $n$ such that $m'$ is a candidate in $t$ it has a bigger timestamp than n. More formally, assume that there exists $k, k' \in O_t$ such that $youngest_s(p) = n_s(k)$ and $n_s(k') = n_s(p)$. Then, either $k \in ov_t(p) \vee k \in be_t(p)$ or there exists $k'' \in O_t$ s.t. $n_s(k'')->ts > n_s(k) ->ts$ and $k'' \in ov_t(p) \vee k'' \in be_t(p)$ and either $k' <^{tr}_s k''$ or $n_s(p) = n_s(k'') \wedge cp_s(p) = R_j \wedge j>3$. 

@@

\end{proof}

>>>>

void push(int x) {
  Node* n = new Node(x,MAX_INT,
                        null,false);
  n->next = pools[myTID];
  pools[myTID] = n;
  int i = TS++;
  n->ts = i;
}
int pop() {
 boolean success = false;
 int maxTS = -1;
 Node* youngest = null;
 while ( !success ) {
   maxTS = -1; youngest = null;
   for(int i=0; i<maxThreads; i++){
     Node* n = pools[i];
     while (n->taken && n->next != n)
       n = n->next;
     if(maxTS < n->ts) {
       maxTS = n->ts; youngest = n;
     }
   }
   if (youngest != null)
     success=CAS(youngest->taken,
                       false,true);
 }
 return youngest->data;
}
\end{lstlisting}
Then, we begin to obtain the LTS ($TSS$) of TS Stack from the algorithm by defining control points and the actions among these control points as seen in Figure X \textcolor{red}{Flow diagram is referenced here}. To simplify the proof, we take the initializations of some local variables together as atomic.

States of the TS-Stack contains the global variables and local variables as fields. Global variables are just elements of their domains and local variables are maps from operation identifiers to their domains. We say $i_q(k)$ for referencing the value of local variable $i$ of operation $k$ in state $q$. There is only one special local variable called $myTID$. Its value is unique to each pending operation in a state i.e., concurrent operations cannot have the same $myTID$ value. TS-Stack states also contains sets $O_a, O_r \in \mathbb{O}$ which are operation identifier sets of push and pops respectively, and the control point function $cp$ which is a map from operation identifiers to the control points set that are presented in the flow diagram Figure X  \textcolor{red}{Flow diagram is referenced here}. Transition relation of the TS-STack is presented in Figure~\ref{fig:transitions:TSSPush} (push rules) and Figure~\ref{fig:transitions:TSSPop} (pop rules).
Next, we show that the linearizability of TS Stack.

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp) \\ 
      d\neq {\tt null}
    }{
      ..., O_a, x, cp,...
      \xrightarrow{inv(push,d,k)} 
     ..., O_a\cup\{k\}, x[k \mapsto d], cp[k \mapsto A_1], ...
    }\hspace{5mm}

    \inferrule[push1]{
      cp(k) = A_1 \\ 
      *n' = (x(k),\texttt{MAX\_INT}, \texttt{null}, \texttt{false})
    }{
      ..., n, cp,...
      \xrightarrow{push1(k)} 
     ..., n[k \mapsto n'], cp[k \mapsto A_2], ...
    }\hspace{5mm}

  \inferrule[push2]{
      cp(k) = A_2 
    }{
      ..., pools, cp,...
      \xrightarrow{push2(k)} 
     ..., pools[myTid(k) \mapsto n(k)], cp[k \mapsto A_3], ...
    }\hspace{5mm}

      \inferrule[push3]{
      cp(k) = A_3 
    }{
      ..., i, TS, cp,...
      \xrightarrow{push3(k)} 
     ..., i[k \mapsto TS], TS+1, cp[k \mapsto A_4], ...
    }\hspace{5mm}

   \inferrule[push4]{
      cp(k) = A_4 \\
      n'(k) = n(k)[ts \mapsto i(k)] \\
      \forall k'. cp(k') = A_6 \implies i(k') < i(k)
    }{
      ..., n, cp,...
      \xrightarrow{push4(k)} 
     ..., n[k \mapsto n'(k)], cp[k \mapsto A_5], ...
    }\hspace{5mm}

   \inferrule[ret-push]{
      cp(k) = A_5 
    }{
      ...,cp,...
      \xrightarrow{ret(push,k)} 
     ...,cp[k \mapsto A_6], ...
    }\hspace{5mm}

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The push derivation rules of $TSS$. We only mention the state components that are modified. Unmentioned state components have the names in the algorithm in the prestate. $*n = (a,b,c,d)$ is shorthand for $n->data = a$, $n->ts = b$, ... $n' = n[ts \mapsto expr]$ is short for $n'->ts = expr$ and all the other fields of $n$ and $n'$ are the same.
  }
  \label{fig:transitions:TSSPush}
\vspace{-6mm}
\end{figure}

\begin{figure} [t]
{\scriptsize
  \centering
  \begin{mathpar}
    \inferrule[call-push]{
      k\not\in dom(cp) \\ 
      d\neq {\tt null}
    }{
      ..., O_a, x, cp,...
      \xrightarrow{inv(push,d,k)} 
     ..., O_a\cup\{k\}, x[k \mapsto d], cp[k \mapsto A_1], ...
    }\hspace{5mm}

      \end{mathpar}
  }
 \vspace{-5mm}
  \caption{The pop derivation rules of $TSS$. We only mention the state components that are modified. Unmentioned state components have the names in the algorithm in the prestate. $*n = (a,b,c,d)$ is shorthand for $n->data = a$, $n->ts = b$, ... $n' = n[ts \mapsto expr]$ is short for $n'->ts = expr$ and all the other fields of $n$ and $n'$ are the same.
  }
  \label{fig:transitions:TSSPop}
\vspace{-6mm}
\end{figure}

\begin{lem}
$TSS$ is a Com(pop)- refinement of $AbsS$. 
\end{lem}
\begin{proof}
Since $AbsS$ is deterministic with respect to $C \cup R \cup Com(pop)$, there exists a Com(pop) forward simulation from $TSS$ to $AbsS$ iff $TSS$ is a Com(pop)-refinement of $AbsS$. Hence, we define the relation $fs$ and show that it is a Com(pop)-forward simulation relation.

@@

\item[\emph{TraverseBefore}] If a pop operation $p$ is currently visiting node $n$ and there is a node $m$ coming before $n$ in the traverse order with a greater timestamp, then the operation that inserts $m$ must be concurrent with $p$. More formally, let $k \in O_t$ such that $n_s(k) <^{tr}_s n_s(p)$ and $n_s(k)->ts \geq n_s(p)->ts$. Then, $k \in ov_t(p)$.
\item[\emph{TraverseAfter}] If a pop operation $p$ is currently visiting node $n$ that is not null and its youngest element $m$ is not null and still not taken in state $s$, then either $m$ is a candidate to be removed by $p$ in $t$ or there exists a later node $m'$ than $n$ such that $m'$ is a candidate in $t$ and it has a bigger timestamp than n. More formally, assume that there exists $k, k' \in O_t$ such that $youngest_s(p)->taken \neq false$, $youngest_s(p) = n_s(k)$ and $n_s(k') = n_s(p)$. Then, either $k \in ov_t(p) \vee k \in be_t(p)$ or there exists $k'' \in O_t$ s.t. $n_s(k'')->ts > n_s(k) ->ts$ and $k'' \in ov_t(p) \vee k'' \in be_t(p)$ and either $k' <^{tr}_s k''$ or $n_s(p) = n_s(k'') \wedge cp_s(p) = R_j \wedge j>3$. 

@@

Next, we will show that $fs$ is really a Com(pop)-forward simulation relation. Except the trivial base case, we case-split on the transition rules. We first assume $(s, \alpha s') \in \delta_C$ and $t \in fs[s]$. Then, we find  corresponding transition $\alpha' \in \Sigma_{AbsS}$ obeying the Com(pop)-forward simulation relation conditions and obtain $t'$ such that $(t, \alpha' t') \in \delta_{AbsS}$  and $t' \in fs[s']$.

We observe that if $\alpha \in C \cup R \cup Com(pop)$, then the corresponding rule in $AbsS$ is $\alpha' = \alpha$. Otherwise, $\alpha' = \epsilon$.

Let the following derivation rule of $TSS$ be the one for describing $\alpha$:
\begin{mathpar}
    \inferrule{
      \psi
    }{
      s
      \xrightarrow{\alpha} 
      s'
    }
\end{mathpar}
and the following one be the derivation rule of $AbsS$ describing $\alpha'$ if $\alpha' \neq \epsilon$ (equivalently $\alpha' = \alpha$):
\begin{mathpar}
    \inferrule{
      \psi'
    }{
      q
      \xrightarrow{\alpha'} 
      q'
    }
\end{mathpar}
For the cases $\alpha' = \alpha$, we first need to show $\alpha'$ is enabled in state $t$ i.e., $t$ satisfies $\psi'$. If this can not be directly obtained from $s$ satisfies $\psi$ and using one or two obvious conditions on $fs$ (since $t \in fs[s]$), we show the derivation in the proof. Then, $t'$ is obtained in a unique way since $AbsS$ is deterministic on its alphabet $ \Sigma_{AbsS} = C \cup R \cup Com(pop)$. The, only other thing to show is $t' \in fs[t']$. We show this by proving that $t'$ does not violate any of the conditions of the $fs$ described above. We only explain why the new instantiations due to the difference between $s'$ and $s$ or the difference between $t'$ and $t$ do not violate the conditions. We skip the instances that we assumed while relating $s$ to $t$.

For the cases in which $\alpha' = \epsilon$, we have $t'=t$ and the only thing to show is $t \in fs[s']$. Again, we only explain why the new instantions due to the difference between $s'$ and $s$ do not violate the conditions.
\item[\textsc{call-push}] The same derivation rule of $TSS$ is applied to $t$ obtain $t'$. The premise of the rule is satisfied by $t$ trivially in the sense explained before. The new vertex $k$ is added to the $O_t$ such that $k$ is maximal, pending and every completed vertex is ordered before $k$ in $t'$. Moreover, $k$ is overlapping with every pending pop. To see that $t' \in fs[s']$ we observe the following: \emph{Nodes} condition is preserved because $k \in O_{t'}$. Since the newly added vertex $k$ is maximal and pending in $t'$, \emph{Pend/Comp} condition is preserved. \emph{Frontiers} and \emph{MaximalOV} conditions are not violated since $k$ is added to $ov(p)$ set for every pending pop operation $p$. 
\item[\textsc{push1}] We have $t' = t$ and show $t \in fs[s']$.\emph{Nodes} and \emph{Pend/Comp} conditions are still satisfied since $k$ remains to be a pending vertex. \emph{TSOrder} is still preserved. Timestamp of $n_{s'}(k)$ is maximal and every other nodes of push operations with maximal timestamp in $s'$ are pending vertices in $t$. Hence there can be no ordering between those pushes and $k$ in $t$ that can violate \emph{TSOrder}. Moreover, $k$ is maximal in $t$ which means that it cannot be ordered before another push $k'$ of which node has a lower timestamp. \emph{TidOrder} is also satisfied. Since $k$ is ordered after every completed push in $t$ and every other push by the same thread is completed, ordering required by the \emph{TidOrder} is present.
\item[\textsc{push2}] We have $t' = t$ and show $t \in fs[s']$. \emph{Nodes} and \emph{Pend/Comp} conditions are still satisfied since $k$ remains to be a pending vertex. One can also see that the \emph{TraverseBefore} condition is preserved. Let the pop $p$ visiting node $m$ and $n_{s'}(k) <^{tr}_{s'} m$. Since $k$ and $p$ are both pending in $s$ and $t \in fs[s]$, $k \in ov_t(p)$ (by the \emph{Frontiers} condition). Hence, \emph{TraverseBefore} is preserved. 
\item[\textsc{push3}] We have $t' = t$ and show $t \in fs[s']$. We consider two cases: $n_s(k)->taken$ is \texttt{true} or it is \texttt{false}. For the former case, $k \notin O_t$. The only new instantiation we check is $k \notin O_t$ does not violate \emph{Nodes} condition while relating $s'$ to $t$. For the latter case, we have $k \in O_t$. \emph{Nodes} and \emph{Pend/Comp} conditions are still satisfied since $k$ remains to be a pending vertex after changing $s$ to $s'$.
\item[\textsc{push4}] We have $t' = t$ and show $t \in fs[s']$. We consider two cases: $n_s(k)->taken$ is \texttt{true} or it is \texttt{false}. For the former case, \emph{Nodes} condition is still satisfied since $k$ remains to be not a vertex. For the latter case \emph{Nodes} and \emph{Pend/Comp} conditions are still satisfied since $k$ remains to be a pending vertex. \emph{TSOrder} condition is still not violated since if $k'<_t k$, then $k'$ is a completed vertex in $s$ and $s'$. By the premise of the rule (which can be shown to hold for every operation at control point $A_4$) $i_s(k') < i_s(k)$ and consequently $n_{s'}(k')->ts < n_{s'}(k)->ts$. Since every other push by the thread of $k$ is completed, \emph{TidOrder} still continues to hold for the same reasons. \emph{TraverseAfter} condition is also preserved. Let $k'$ be the push and $p$ be the pop such that $n_s(k') = youngest_s(p)$, $n_s(k') \leq^{tr}_s n_s(k)$, $n_s(k')->ts < n_s(k)->ts$ and $k \in ov_t(p)$ or $k \in be_t(p)$. Assume $n_{s'}(k')->ts \geq N_{s'}(k)->ts$ after the action. Then, $k'$ must be a pending push both in $s$ and $s'$ by the premise of the derivation rule and $k' \in ov_t(p)$ must be true by \emph{Frontiers} condition and $t \in fs[s]$. Hence, the \emph{TraverseAfter} condition is preserved.
\item[\textsc{ret-push}] We consider two cases, $n_s(k)->taken$ is \texttt{false} or \texttt{true}. For the former case, we obtain $t'$ by applying \textsc{ret-push1} rule of $AbsS$. \emph{Nodes} and \emph{Pend/Comp} conditions are still satisfied since $k$ becomes a completed vertex in $t'$. \emph{Frontiers} condition still holds since although $k$ become a maximally closed vertex in $t'$, we have $k \in ov_{t'}(p)$ for all pending nodes $p$ (due to \emph{Frontiers} condition, $t \in fs[s]$ and $k$ was a pending operation in state $t$, $k \in ov_t(p)$).

For the latter case, we obtain $t'$ by applying \textsc{ret-push2} rule of $AbsS$. \emph{Nodes} condition is still satisfied since $k \notin O_{t'}$. 
\end{itemize}
\end{proof}
